---
title: "Non-interactive fitting with fitSpectrum"
author: "Rahul Dey & Hannes Becher"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Non-interactive fitting with fitSpectrum}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(Tetmer)
```

## Overview

`fitSpectrum()` runs the same optimisation machinery as the Autofit
mode of the interactive `tetmer()` app, but as a plain function call
that returns a `tetmerFit` object directly. This makes it possible to
fit many spectra in a script or loop, on a cluster, or in an automated
pipeline, without needing a Shiny session or the `shiny` package at
all.

## Fitting a diploid spectrum

The built-in `E030` spectrum is from a diploid plant. Its dominant
peak sits at a k-mer multiplicity of around 108, so a reasonable
starting range for the k-mer coverage is somewhat below that value:

```{r}
fit_d <- fitSpectrum(E030,
                      model  = "d",
                      kcov   = c(30, 60),
                      vf     = c(1, 10),
                      theta  = c(-2, 0.6),
                      gs     = c(6, 9),
                      xrange = c(45, 150))
fit_d
```

The arguments mirror the sliders in the interactive app's Autofit
panels: `kcov`, `vf`, `theta`, and `gs` each take a two-element vector
giving the lower and upper bound to search within, and `xrange` gives
the multiplicity range of the data to fit over -- which should exclude
any contamination peak at low multiplicity, just as in the interactive
app.

To visualise the fitted curve with `plot.spectrum()`, attach the fit to
the original spectrum and then plot that spectrum:

```{r, fig.width = 7, fig.height = 5}
spec_d <- addFit(E030, fit_d)
plot(spec_d)
```

## Fitting a polyploid spectrum

The built-in `E028` spectrum is from an allotetraploid plant. Fitting
an allopolyploid model additionally requires a `diverg` range for the
divergence time between sub-genomes (in units of 2*Ne*), which defaults
to `c(0.1, 100)` if not supplied:

```{r}
fit_tal <- fitSpectrum(E028,
                        model  = "tal",
                        kcov   = c(37, 45),
                        vf     = c(1, 10),
                        theta  = c(-2, 0.6),
                        gs     = c(6, 9),
                        xrange = c(25, 200),
                        diverg = c(0.1, 100))
fit_tal
```

The same plotting workflow works for polyploid fits:

```{r, fig.width = 7, fig.height = 5}
spec_tal <- addFit(E028, fit_tal)
plot(spec_tal)
```

## Available models

The `model` argument accepts one of:

- `"d"` -- diploid
- `"traaa"` -- autotriploid
- `"traab"` -- allotriploid
- `"tau"` -- autotetraploid
- `"tal"` -- allotetraploid
- `"tse"` -- segmental allotetraploid, which additionally requires a
  `pallo` range (proportion of the genome that is allopolyploid),
  defaulting to `c(0.01, 0.99)`

These correspond to the same model choices offered in the interactive
app's "Model" panel.

## Handling failed optimisation

Optimisation is run with random restarts, controlled by `maxAttempts`
(default `5`). If every attempt fails to converge, `fitSpectrum()`
issues a warning and returns `NULL` rather than a `tetmerFit` object.
This can happen, for instance, if `xrange` extends beyond the
multiplicity values present in the spectrum:

```{r}
failed_fit <- fitSpectrum(E030,
                           model  = "d",
                           kcov   = c(50, 200),
                           vf     = c(1, 10),
                           theta  = c(-2, 0.6),
                           gs     = c(6, 9),
                           xrange = c(45, 5000),
                           maxAttempts = 2)
is.null(failed_fit)
```

In practice, a `NULL` result usually means the parameter ranges are
implausible for the data -- as here, where `xrange` reaches far beyond
the maximum multiplicity actually present in `E030`. Correcting the
ranges, or increasing `maxAttempts` to allow more random restarts, is
generally enough to resolve it.

## Batch fitting

Because `fitSpectrum()` is a plain function, it fits naturally into a
loop over many samples, for example when each sample's spectrum has
been saved to its own file with `read.spectrum()`:

```{r, eval = FALSE}
files <- list.files("spectra/", full.names = TRUE)

fits <- lapply(files, function(f) {
  spec <- read.spectrum(f, nam = tools::file_path_sans_ext(basename(f)), k = 21)
  fitSpectrum(spec,
              model  = "d",
              kcov   = c(5, 200),
              vf     = c(1, 10),
              theta  = c(-2, 0.6),
              gs     = c(6, 9),
              xrange = c(45, 250))
})
names(fits) <- tools::file_path_sans_ext(basename(files))
```

Each element of `fits` is either a `tetmerFit` object or `NULL` if
optimisation failed for that sample, so it is worth checking for `NULL`
before summarising results across the batch.

For working with the `spectrum` and `tetmerFit` objects returned here
-- attaching a fit to its spectrum, or exporting the result to a file
-- see the "The spectrum and tetmerFit classes" vignette.
