---
title: "The spectrum and tetmerFit classes"
author: "Hannes Becher"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The spectrum and tetmerFit classes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(Tetmer)
```

## Overview

Tetmer represents k-mer spectra and model fits with two S4 classes:

- `spectrum` holds a named k-mer spectrum, together with any fits that
  have been computed for it.
- `tetmerFit` holds the result of a single model fit: the model type,
  fitted parameters, the ranges used in the optimisation, and some
  convergence information.

This vignette introduces both classes, shows how a `tetmerFit` is
attached to a `spectrum` with `addFit()`, and demonstrates the
human-readable export and re-import functions `write.spectrum()` and
`read.spectrum()`.

## The spectrum class

`spectrum` objects have four slots: `name`, `data`, `k`, and `fits`.
The built-in `E030` dataset is one example:

```{r}
E030
slotNames(E030)
```

The `data` slot is a plain `data.frame` with two columns, `mult` (k-mer
multiplicity) and `count` (the number of distinct k-mers observed at
that multiplicity):

```{r}
head(E030@data)
```

`k` records the k-mer length used to generate the spectrum, and `fits`
is a list of `tetmerFit` objects -- empty until a fit has been added.

A `plot` method is defined for `spectrum` objects, which is often
easier to work with on a log-log scale to see the full range of the
data:

```{r, fig.width = 6, fig.height = 4}
plot(E030, log = "xy")
```

## The tetmerFit class

A `tetmerFit` object is produced by `fitSpectrum()` (see the
"Non-interactive fitting with fitSpectrum" vignette) or by clicking
**Done** after an Autofit run in the interactive app. It records eight
slots: `model`, `par`, `ranges`, `xrange`, `convergence`, `value`, `k`,
and `version`.

```{r}
fit <- fitSpectrum(E030,
                    model  = "d",
                    kcov   = c(50, 200),
                    vf     = c(1, 10),
                    theta  = c(-2, 0.6),
                    gs     = c(6, 9),
                    xrange = c(45, 250))
fit
slotNames(fit)
fit@par
```

`par` holds the fitted parameter values, `ranges` records the bounds
that were searched, `xrange` records the multiplicity range the fit
was computed over, and `convergence`/`value` come directly from the
underlying call to `optim()`. `version` records the Tetmer version
used, which is useful when comparing fits produced at different times.

## Attaching a fit to its spectrum: addFit()

`addFit()` appends a `tetmerFit` to a `spectrum`'s `fits` slot,
returning an updated `spectrum` object (it does not modify anything in
place, in keeping with R's usual copy-on-modify semantics):

```{r}
spec_with_fit <- addFit(E030, fit)
spec_with_fit
```

Multiple fits can be attached to the same spectrum -- for example, to
keep a record of fits obtained under different models or parameter
ranges for comparison:

```{r}
fit_narrower <- fitSpectrum(E030,
                             model  = "d",
                             kcov   = c(90, 120),
                             vf     = c(1, 10),
                             theta  = c(-2, 0.6),
                             gs     = c(6, 9),
                             xrange = c(45, 250))
spec_with_fits <- addFit(spec_with_fit, fit_narrower)
spec_with_fits
```

## Exporting with write.spectrum()

`write.spectrum()` writes a `spectrum` object to a plain text file.
The spectrum data are written as two columns, preceded by a block of
`#`-prefixed comment lines recording the spectrum's name and k-mer
length, and -- for each stored fit -- its model, fitted parameters,
parameter ranges, and convergence information:

```{r}
tf <- tempfile(fileext = ".txt")
write.spectrum(spec_with_fit, tf)
cat(readLines(tf, n = 15), sep = "\n")
```

Because the metadata lines start with `#`, the file remains a valid
plain-text k-mer spectrum: it can be inspected with any text editor or
scrolled through with `less`, and it can be read by any tool that
expects a two-column, `#`-commented spectrum file, including Tetmer's
own `read.spectrum()`.

## Re-importing with read.spectrum()

`read.spectrum()` reads the multiplicity/count data back in as a new
`spectrum` object:

```{r}
reloaded <- read.spectrum(tf, nam = "E030 reloaded", k = 21)
reloaded
all.equal(reloaded@data, spec_with_fit@data, check.attributes = FALSE)
```

Note that `read.spectrum()` restores the spectrum data itself, but not
the fit records: the `# fit.*` metadata lines written by
`write.spectrum()` are for human inspection and downstream reference
rather than being parsed back in, so `reloaded@fits` is an empty list
even though the file on disk records one fit. If you need to keep
working with a fit programmatically, keep a reference to the
`tetmerFit` object (or the `spectrum` it was added to) in your R
session, rather than relying on re-reading the exported file.

## Putting it together

A typical scripted workflow combines all of the pieces shown above:
read or select a spectrum, fit a model to it, attach the fit, and
export the annotated spectrum for later reference:

```{r, eval = FALSE}
spec <- read.spectrum("path/to/mySpectrumFile", nam = "My sample", k = 21)
fit  <- fitSpectrum(spec,
                     model  = "d",
                     kcov   = c(5, 200),
                     vf     = c(1, 10),
                     theta  = c(-2, 0.6),
                     gs     = c(6, 9),
                     xrange = c(45, 250))
spec <- addFit(spec, fit)
write.spectrum(spec, "mySample_fitted.txt")
```
