---
title: "The spectrum and tetmerFit classes"
author: "Hannes Becher"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The spectrum and tetmerFit classes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(Tetmer)
```

## Overview

Tetmer represents k-mer spectra and model fits with two S4 classes:

- `spectrum` holds a named k-mer spectrum, together with any fits that
  have been computed for it.
- `tetmerFit` holds the result of a single model fit: the model type,
  fitted parameters, the ranges used in the optimisation, and some
  convergence information.

This vignette introduces both classes, shows how a `tetmerFit` is
attached to a `spectrum` with `addFit()`, and demonstrates the
human-readable export and re-import functions `write.spectrum()` and
`read.spectrum()`.

## The spectrum class

`spectrum` objects have four slots: `name`, `data`, `k`, and `fits`.
The built-in `E030` dataset is one example:

```{r}
E030
slotNames(E030)
```

The `data` slot is a plain `data.frame` with two columns, `mult` (k-mer
multiplicity) and `count` (the number of distinct k-mers observed at
that multiplicity):

```{r}
head(E030@data)
```

`k` records the k-mer length used to generate the spectrum, and `fits`
is a list of `tetmerFit` objects -- empty until a fit has been added.

A `plot` method is defined for `spectrum` objects, which is often
easier to work with on a log-log scale to see the full range of the
data:

```{r, fig.width = 6, fig.height = 4}
plot(E030, log = "xy")
```

## The tetmerFit class

A `tetmerFit` object is produced by `fitSpectrum()` (see the
"Non-interactive fitting with fitSpectrum" vignette) or by clicking
**Done** in the interactive app. It records nine slots: `fitType`,
`model`, `par`, `ranges`, `xrange`, `convergence`, `value`, `k`, and
`version`.

```{r}
fit <- fitSpectrum(E030,
                    model  = "d",
                    kcov   = c(50, 200),
                    vf     = c(1, 10),
                    log10theta = c(-2, 0.6),
                    log10Mbp   = c(0, 3),
                    xrange = c(45, 250))
fit
slotNames(fit)
fit@par
```

`fitType` records whether the fit came from Autofit (`"auto"`) or from
the current Manual settings (`"manual"`). `par` holds the fitted
parameter values, `ranges` records the bounds that were searched for
auto fits, `xrange` records the multiplicity range the fit was computed
over, and `convergence`/`value` come directly from the underlying call to
`optim()` for auto fits. Theta ranges are stored as `log10theta`, and
genome-size ranges are stored as `log10Mbp`, the log10 of the monoploid
non-repetitive genome size in Mbp. `version` records the Tetmer version
used, which is useful when comparing fits produced at different times.

## Attaching a fit to its spectrum: addFit()

`addFit()` appends a `tetmerFit` to a `spectrum`'s `fits` slot,
returning an updated `spectrum` object (it does not modify anything in
place, in keeping with R's usual copy-on-modify semantics):

```{r}
spec_with_fit <- addFit(E030, fit)
spec_with_fit
```

Multiple fits can be attached to the same spectrum -- for example, to
keep a record of fits obtained under different models or parameter
ranges for comparison:

```{r}
fit_narrower <- fitSpectrum(E030,
                             model  = "d",
                             kcov   = c(90, 120),
                             vf     = c(1, 10),
                             log10theta = c(-2, 0.6),
                             log10Mbp   = c(0, 3),
                             xrange = c(45, 250))
spec_with_fits <- addFit(spec_with_fit, fit_narrower)
spec_with_fits
```

## Exporting with write.spectrum()

`write.spectrum()` writes a `spectrum` object to a plain text file.
The spectrum data are written as two columns, preceded by a block of
`#TETMER`-prefixed comment lines recording the spectrum's name and
k-mer length, and -- for each stored fit -- its model, fitted
parameters, parameter ranges, and convergence information:

```{r}
tf <- tempfile(fileext = ".txt")
write.spectrum(spec_with_fit, tf)
cat(readLines(tf, n = 15), sep = "\n")
```

Because the metadata lines start with `#`, the file remains a valid
plain-text k-mer spectrum: it can be inspected with any text editor or
scrolled through with `less`, and it can be read by tools that expect a
two-column, `#`-commented spectrum file.

## Re-importing with read.spectrum()

`read.spectrum()` reads the multiplicity/count data back in as a new
`spectrum` object:

```{r}
reloaded <- read.spectrum(tf, nam = "E030 reloaded", k = 21)
reloaded
all.equal(reloaded@data, spec_with_fit@data, check.attributes = FALSE)
```

For files written by `write.spectrum()`, `read.spectrum()` also restores
the `#TETMER` metadata, including any stored fit records. Explicit
`nam =` and `k =` arguments still override the name and k-mer length
recorded in the file.

## Tetmer spectrum text format

Tetmer spectrum files are plain text. Lines beginning with `#TETMER `
are Tetmer metadata. Other lines beginning with `#` are ordinary
comments, which makes the files compatible with tools that expect
commented two-column k-mer spectra. The spectrum data follow the
metadata as two whitespace-separated columns:

- `mult`: k-mer multiplicity
- `count`: k-mer count

### Format 1

The current text format is identified by:

```text
#TETMER format: 1
```

The following metadata fields are written once per spectrum:

```text
#TETMER tetmer.version: <character>
#TETMER name: <character>
#TETMER k: <number>
#TETMER fits: <integer>
```

The `tetmer.version` field records the Tetmer package version that wrote
the file. The `format` field records the text-file format version and is
the field readers should use to decide how to parse the file.

For each stored fit `i`, where `i` runs from `1` to `fits`, Tetmer writes
these fields:

```text
#TETMER fit.i.fitType: <manual|auto>
#TETMER fit.i.model: <character>
#TETMER fit.i.k: <number>
#TETMER fit.i.version: <character>
#TETMER fit.i.convergence: <number>
#TETMER fit.i.value: <number>
#TETMER fit.i.xrange: <number> <number>
```

Fitted parameters and optimisation ranges are written with one line per
named entry:

```text
#TETMER fit.i.par.<name>: <number>
#TETMER fit.i.range.<name>: <number> <number>
```

Theta optimisation ranges use the name `log10theta`; genome-size
optimisation ranges use the name `log10Mbp`.

After the metadata and any ordinary comment lines, the spectrum itself is
stored as whitespace-separated data:

```text
# mult count
1 123
2 456
```

## Putting it together

A typical scripted workflow combines all of the pieces shown above:
read or select a spectrum, fit a model to it, attach the fit, and
export the annotated spectrum for later reference:

```{r, eval = FALSE}
spec <- read.spectrum("path/to/mySpectrumFile", nam = "My sample", k = 21)
fit  <- fitSpectrum(spec,
                     model  = "d",
                     kcov   = c(5, 200),
                     vf     = c(1, 10),
                     log10theta = c(-2, 0.6),
                     log10Mbp   = c(0, 3),
                     xrange = c(45, 250))
spec <- addFit(spec, fit)
write.spectrum(spec, "mySample_fitted.txt")
```
